# Week 1

## Student exploration

1. Select one of the papers from the list below

> Selected Paper: [Steyerberg et al. (2001)](https://www.sciencedirect.com/science/article/pii/S0895435601003419) - Internal validation of predictive models: efficiency of some procedures for logistic regression analysis

2. For the initial exploration, do the following:
   - Prepare a summary of the key ideas. Be prepared to describe/illustrate the concepts in a concrete example.
   - Review the instructions for the student presentation and come with an idea/sketch of the simulation you'll perform.

> **Key Ideas**
>
> 1. From the opening line of the abstract: *"The performance of a predictive model is overestimated when simply determined on the sample of subjects that was used to construct the model."* <br>
> The authors expand on this idea of optimism. When the model is trained and evaluated on only data within the sample, the estimated performance will be optimistic compared with the real performance on new data that is completely outside the sample. This difference is between apparent performance and test performance.
>
> 2. From the closing paragraph of the introduction: *"In this study we compare the efficiency of internal validation procedures for predictive logistic regression models."* (also basically the title lol) <br>
> The authors walked through a comparison of the split-sample approach (test-train splits), cross-validation, and bootstrapping. Each one is an internal validation procedure that is commonly used for evaluating models.
>
> 3. From the opening line of section 2.4: "We evaluated several internal validation procedures that aim to test performance more accurately than is achieved by the apparent performance estimate." <br>
> The authors set out to determine which validation procedure of those listed above is best for the logistic regression model in question. They used the GUSTO-I dataset, detailed below, and fit multiple regression models for prediction.
> 
> **Dataset:** GUSTO-I; n=40,830; 2,851 deaths <br>
> Working with a dataset of 30-day mortality after an acute myocardial infarction.
> What does that mean? They're looking at the % of patients who die from any cause within 30 days of a heart attack.
>
> **What they want us to do:**
> Stop doing CV and split-sample (test-train splits), and start doing regular bootstraping instead for estimation of accuracy!
> 
> **How they showed it:** Logistic regression was fitted in each sample with a previously specified set of eight dichotomous predictors (including shock, age>65, diabetes, etc).
> The apparent performance of the model fitted to the full GUSTO-I data was the reference point from which they compared the other performance measures. *"The large size of the total GUSTO-I data set makes sure that the optimism in this performance estimate is negligible."*
> As mentioned above, they compared variants of split-sample, cross-validation, and bootstrap (including the .632 and .632+ methods) specifically at EPV=10 to analyze estimates of model performance.
>
> **Some results:**
> 
> - The optimism and the variability of the model performance estimates reduced as sample size increased.
> - *"The regular bootstrap resulted in the most accurate estimates of model performance: the bias was close to zero and the variability similar to that of the apparent performance."*
> - Bootstrap resampling resulted in stable and nearly unbiased estimates of performance.
> - Relation to some other papers: *"a puzzling observation was that the .632 and the .632+ bootstrap variants were not superior to the regular bootstrap."*
> 
> **Definitions** <br>
>
> - *Events Per Variable (EPV)*: The number of outcome events (i.e. number of deaths) divided by the number of predictor variables in the model. Recommended to be at least 10 to provide an adequate predictive model. <br>
> - *Internal Validation*: All methods of assessing performance using only the dataset you were provided. <br>
> - *External Validation*: Assessing performance using a TRULY NEW dataset. Something from another clinic, a separate experiment, real world data, etc. <br>
> - *Apparent Performance*: Predictive performance estimated on the training sample. Optimistically biased, because the same data is used to fit and evaluate the model. <br>
> - *Test Performance*: Predictive performance estimated on the test data (held-out sample), an independent test set consisting of all patients from GUSTO-I except the patients in the random sample. <br>
> - *Optimism*: The difference between apparent performance and a model's true performance on new data. <br>
> - *Split-sample approach*: Randomly split training data in two parts: one to develop the model and another to measure its performance. <br>
> - *Cross-validation*: Model is developed on one part of the data and tested on the other part, while these samples are randomly taken repeatedly and the average is taken as an estimate of the performance. <br>
> - *Bootstrap*: Replicates the process of sample generation from an underlying population by drawing samples with replacement from the original data set, of the same size as the original data set. <br>
> - *Discrimination*: The ability to distinguish high-risk subjects from low-risk subjects. Commonly quantified by a measure of concordance, the *c* statistic. <br>
> - *Calibration*: Whether the predicted probabilities agree with the observed probabilities. Well-calibrated models have a slope of 1, while models providing too extreme predictions have a slope less than 1. Well-calibrated models also have an intercept of 0, while models providing predictions that are too high or too low on average have an intercept different from 0.

> **Sketch of the simulation**: Replicate a simulation from the paper 
> 
> The goal is to demonstrate that the regular bootstrap returns the optimal estimates of internal validity.
>
> - Start with the GUSTO-I dataset (from R or Python package)
> - Generate sample and independent test sets for EPV = 10
> - Complete split sample and regular boostrap on the sample dataset
> - Compare Predictive performance (*c* statistic, calibration slope, Brier score, D(=scaled$\chi^2$), $R^2$)
> 

3. For the student presentation, do one of the following and come prepared to present it to the class:
   - **Replicate a simulation or example from the paper** [SELECTION]
   - Develop your own simulation study which explores a concept discussed in the paper
   - Complete the simulation study described below.

> **Replicating the example from the paper**
> 
> The following sections replicating the simulation were assisted by Claude. I am still working to develop a full understanding of the code, formulas, and methods completed within. Formulas and comments will be added progressively.

> **Dataset:** GUSTO-I; n=40,830; 2,851 deaths <br>
> Available in full anonymized version via `predtools` package in R.

> **Figures:** Replicating a version of Figure 2 from the paper
>
> In the paper, figure 2 shows bias and variability for a few of the procedures for estimating model performance in independent data for EPV = 10, including apparent performance, split-sample validation on 50% or 33% of the sample, cross-validation on 50% or 10% of the sample, and regular bootstrapping. 
> In this simulation, I will show apparent performance, split-sample validation on 33% of the sample, and regular bootstrapping. I will also only show metrics for calibration and discrimination (*c* statistic, calibration slope, Brier score).

> **Methods:** Replicating the steps of the paper

> 1. Load in the dataset
> 2. Define logistic regression model with 8 predictors from the paper
> 3. Generate a sub-sample for EPV = 10 (set a seed to keep results steady)
> 4. [Not needed for this simulation] Use model to compute the benchmark performance on the full dataset
> 5. Use model to compute apparent performance on sub-sample dataset [simulated many times to get bias and variability]
> 6. Use model to compute split-sample performance on sub-sample dataset [""]
> 7. Use model to compute bootstrap performance on sub-sample dataset [""]
> 8. Build figure 2 to compare performance!

```{r setup}

## Loading in Gusto-I dataset from predtools

#install.packages("predtools")
library(predtools)
data(gusto)

str(gusto)
```

> From the paper: *"A logistic regression model was fitted in each sample consisting of a previously specified set of eight dichotomous predictors: shock, age>65 years, high risk (anterior infarct location or previous MI), diabetes, hypotension (systolic blood pressure <100 mmHg), tachycardia (pulse>80), relief of chest pain>1 hr, female gender"*

> Variables we want are:
> 
> - Outcome: `day30`
> - Predictors:
> - Shock: `sho`
> - Age > 65 years: `age` (needs transform to binary)
> - High Risk: `ant` or `pmi` (needs transform to binary)
> - Diabetes: `dia`
> - Hypotension: `sysbp` (needs transform to binary)
> - Tachycardia: `pulse` (needs transform to binary)
> - Relief of chest pain > 1hr: `ttr`
> - Female gender: `sex`

```{r transform}

gusto_df <- data.frame(
   day30 = as.integer(gusto$day30),
   shock = as.integer(gusto$sho),
   age65 = as.integer(gusto$age > 65),
   highrisk = as.integer(gusto$ant == 1 | gusto$pmi == "yes"),
   diabetes = as.integer(gusto$dia),
   hypotension = as.integer(gusto$sysbp < 100),
   tachycardia = as.integer(gusto$pulse > 80),
   ttr = as.integer(gusto$ttr),
   female = as.integer(gusto$sex == "female")
)

head(gusto_df)

```

> All variables are now transformed to match the "eight dichotomous predictors" from the paper. Below, the logistic regression model is defined in R code for the upcoming simulations.

```{r model definition}

model_formula <- day30 ~ shock + age65 + highrisk + diabetes + hypotension + tachycardia + ttr + female

```

> From the paper: *"For example, for EPV=10 and 7.0% mortality, all training and test samples consisted of 1145 and 39,685 patients, with 80 and 2771 deaths, respectively."*
```{r sample}

## set seed for reproducibility
set.seed(17)

## Creating ID column for tracking which patients are in what samples
gusto_df$row_id <- seq_len(nrow(gusto_df))

## Function to draw the training sample for each time this needs to be run
draw_dev_sample <- function(data, sample_total = 1145, target_events = 80){
   
   ## EPV = 10 with 8 predictors requires 80 events (deaths)
   target_events <- target_events
   target_non_events <- sample_total-target_events

   ## Stratified sampling to ensure desired number of events
   # Split into events and non-events
   events_pool     <- gusto_df[gusto_df$day30 == 1, ]
   nonevents_pool  <- gusto_df[gusto_df$day30 == 0, ]

   # Draw the stratified subsample
   dev_events    <- events_pool[sample(nrow(events_pool), target_events), ]
   dev_nonevents <- nonevents_pool[sample(nrow(nonevents_pool), target_non_events), ]

   dev_sample <- rbind(dev_events, dev_nonevents)

   # Shuffle row order so events aren't grouped together
   dev_sample <- dev_sample[sample(nrow(dev_sample)), ]
}

## Function to draw the testing sample (all the rows that are not in the training sample)
draw_test_set <- function(data, dev_sample) {
  data[!(data$row_id %in% dev_sample$row_id), ]
}

```

> Building reusable functions for computing desired metrics:
> 
> **Metric definitions**
>
> **C-statistic (discrimination)** — the probability that a randomly chosen patient who died is assigned a higher predicted risk than a randomly chosen patient who survived:
> $$c = P(\hat{p}_i > \hat{p}_j \mid Y_i = 1,\ Y_j = 0)$$

> Estimated empirically as the proportion of all (event, non-event) pairs correctly ordered by the model, with ties counted as half-correct:
> $$\hat{c} = \frac{1}{n_1 n_0} \sum_{i:\, Y_i=1} \sum_{j:\, Y_j=0} \left[ \mathbb{1}(\hat{p}_i > \hat{p}_j) + 0.5 \cdot \mathbb{1}(\hat{p}_i = \hat{p}_j) \right]$$

> where $n_1$ is the number of events and $n_0$ is the number of non-events.
> In this case from the paper, since we have a binary outcome, c is identical to the area under the receiver operating characteristic (ROC) curve; c varies between 0.5 and 1.0 for sensible models (the higher the better).
>
> **Calibration slope** — the coefficient $\beta$ from an auxiliary logistic regression of the observed outcome on the linear predictor (logit of the model's predicted probability):
> $$\text{logit}\big(P(Y=1)\big) = \alpha + \beta \cdot \text{linear predictor}$$

> A well-calibrated model has $\beta = 1$. $\beta < 1$ indicates predictions that are too extreme (overfitting); $\beta > 1$ indicates predictions that are too conservative.
>
> **Brier score** — mean squared error between predicted probability and observed outcome:
> $$\text{Brier} = \frac{1}{n} \sum_{i=1}^{n} (\hat{p}_i - Y_i)^2$$

> Ranges from 0 (perfect) to 0.25 (worthless) for a well-behaved binary outcome model.
> Exactly the same formula as MSE, but specifically used for binary prediction models.

```{r metrics}

library(pROC)

compute_metrics <- function(y, pred) {
  # Guard against predicted probabilities of exactly 0 or 1,
  # which would break the logit transform below
  eps <- 1e-10
  pred <- pmin(pmax(pred, eps), 1 - eps)

  # Discrimination: c-statistic (AUC)
  c_stat <- as.numeric(pROC::auc(y, pred, quiet = TRUE))

  # Calibration slope: regress outcome on the linear predictor (logit of pred)
  lp <- log(pred / (1 - pred))
  cal_fit <- glm(y ~ lp, family = binomial())
  cal_slope <- unname(coef(cal_fit)["lp"])

  # Brier score
  brier <- mean((y - pred)^2)

  c(c_stat = c_stat, cal_slope = cal_slope, brier = brier)
}


```

> Computing apparent performance for the logistic regression model on the training sample dataset
```{r apparent}

get_apparent <- function(sample_data, formula) {
  fit <- glm(formula, data = sample_data, family = binomial())
  pred <- predict(fit, type = "response")
  list(fit = fit, metrics = compute_metrics(sample_data$day30, pred))
}

```

> Defining the function to evaluate a given model on new data for the test performance
```{r evaluate}
eval_on_data <- function(fit, newdata) {
  pred <- predict(fit, newdata = newdata, type = "response")
  compute_metrics(newdata$day30, pred)
}
```

> Function for the split-sample validation, essentially running a test-train split on the development sample and fitting the model.
```{r split-sample}

split_sample_validate <- function(sample_data, formula, train_frac = 0.6667) {
  n_train <- round(nrow(sample_data) * train_frac)
  train_idx <- sample(nrow(sample_data), n_train)

  train_set <- sample_data[train_idx, ]
  test_set  <- sample_data[-train_idx, ]

  fit <- glm(formula, data = train_set, family = binomial())
  eval_on_data(fit, test_set)
}

```

> Defining function for the regular bootstrap method. Setting B=100 as the authors did in the paper.
> Sampling from the sample distribution with replacement, then training on that sample.
> Computing performance in the following way:
> *The performance in the bootstrap sample represents estimation of the apparent performance, and the performance in the original sample represents test performance. The difference between these performances is an estimate of the optimism in the apparent performance. This difference is averaged to obtain a stable estimate of the optimism. The optimism is subtracted from the apparent performance to estimate the internally validated performance [7]: estimated performance = apparent performance − average(bootstrap performance − test performance).*
> In other words, we evaluate $\text{model}_b^*$ on the bootstrap sample itself (its own "apparent" performance), and on the original sample (its "test" performance).
> $$\text{optimism}_b = \text{Perf}(\text{model}_b^*,\ \text{sample}_b^*) \;-\; \text{Perf}(\text{model}_b^*,\ \text{sample}_{\text{orig}})$$
> Average the optimism across all $B$ replicates:

> $$\overline{\text{optimism}} = \frac{1}{B}\sum_{b=1}^{B} \text{optimism}_b$$

> Subtract this from the apparent performance of the model fitted on the original (development) sample to get the bias-corrected estimate:

> $$\text{estimated performance} = \text{apparent performance} - \overline{\text{optimism}}$$

```{r bootstrap}

bootstrap_correct <- function(sample_data, formula, apparent_metrics, B = 100) {
  n <- nrow(sample_data)
  optimism_sum <- c(c_stat = 0, cal_slope = 0, brier = 0)

  for (b in 1:B) {
    boot_idx <- sample(n, n, replace = TRUE)
    boot_sample <- sample_data[boot_idx, ]

    boot_fit <- glm(formula, data = boot_sample, family = binomial())

    boot_apparent <- compute_metrics(
      boot_sample$day30,
      predict(boot_fit, type = "response")
    )
    boot_test <- compute_metrics(
      sample_data$day30,
      predict(boot_fit, newdata = sample_data, type = "response")
    )

    optimism_sum <- optimism_sum + (boot_apparent - boot_test)
  }

  avg_optimism <- optimism_sum / B
  apparent_metrics - avg_optimism
}

```

> 
> Function 7: run one full repetition (ties everything together)

```{r fn-one-repetition}
# This is the "outer loop body" -- one full pass of the pipeline,
# matching one box in each boxplot of Figures 1 and 2.
run_one_repetition <- function(data, formula, sample_total = 1145, target_events = 80, B = 100) {

  dev_sample <- draw_dev_sample(data, sample_total, target_events)
  test_set   <- draw_test_set(data, dev_sample)

  # Apparent performance + the fitted model (reused below)
  app <- get_apparent(dev_sample, formula)
  apparent_metrics <- app$metrics

  # Test performance: evaluate the dev_sample-fitted model on truly new patients
  test_metrics <- eval_on_data(app$fit, test_set)

  # Split-sample (50/50, matching your simplified scope)
  split_metrics <- split_sample_validate(dev_sample, formula, train_frac = 0.6667)

  # Bootstrap-corrected
  boot_metrics <- bootstrap_correct(dev_sample, formula, apparent_metrics, B = B)

  # MATH NOTE: delta = method's estimate - test performance.
  # delta = 0 means the method perfectly recovered true generalization
  # performance for this repetition. This is exactly Figure 2's y-axis.
  data.frame(
    method = rep(c("apparent", "split_sample", "bootstrap"), each = 3),
    metric = rep(c("c_stat", "cal_slope", "brier"), times = 3),
    delta  = c(
      apparent_metrics - test_metrics,
      split_metrics - test_metrics,
      boot_metrics - test_metrics
    )
  )
}
```

> Run the simulation loop [decreased to 10 for faster preview]

```{r run-simulation}
# NOTE: reduced from the paper's 500 repetitions to keep runtime
# reasonable for a live presentation. State this explicitly when
# presenting -- it's a deliberate, documented scope reduction, not
# an error.
n_reps <- 10

set.seed(17)
results_list <- vector("list", n_reps)

for (i in seq_len(n_reps)) {
  results_list[[i]] <- run_one_repetition(
    data = gusto_df,
    formula = model_formula,
    sample_total = 1145,
    target_events = 80,
    B = 100
  )
}

results <- do.call(rbind, results_list)
```

> Visualize: recreate Figure 2

```{r plot-figure2}
library(ggplot2)

ggplot(results, aes(x = method, y = delta)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title = "Bias of internal validation methods at EPV = 10",
    subtitle = paste0(n_reps, " repetitions (paper used 500)"),
    x = "Method",
    y = "Delta (estimated - test performance)"
  ) +
  theme_minimal()
```

> Summarize bias and RMSE (reproduces Table 2's structure)

```{r summarize-table2}
# MATH NOTE:
#   Bias = mean(delta)                -- average signed error
#   RMSE = sqrt(mean(delta^2))        -- root mean SQUARED error,
#                                          NOT centered around the bias.
#          RMSE^2 = Bias^2 + Var(delta)
library(dplyr)

summary_table <- results %>%
  group_by(method, metric) %>%
  summarise(
    bias = mean(delta),
    rmse = sqrt(mean(delta^2)),
    .groups = "drop"
  )

summary_table
```

<!--
### Extra notes for later

```{r}

library(ggplot2)

y <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
yhat <- c(1, 2.5, 4, 3.4, 5, 6, 6.5, 10, 9, 11)

df <- data.frame(predicted = yhat, actual = y)

ggplot(df, aes(x = predicted, y = actual)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(x = "Predicted Values (yhat)",
       y = "Actual Values (y)") +
  theme_minimal()

```
-->

<!--
```{python}

2+2

```

-->

<!--
```{r benchmark-model}
benchmark_fit <- glm(model_formula, data = gusto_df, family = binomial())

benchmark_pred <- predict(benchmark_fit, type = "response")

```
-->
